---
title: "Analysis of PARP1 mRNA levels in scRNA seq"
author: "Yizhou Yu"
date: 'updated: <i>`r format(Sys.Date(),''%b-%d-%Y'')`</i></h4>'
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
    keep_md: yes
---
download data: 

`curl -o local.rds "https://corpora-data-prod.s3.amazonaws.com/f541adea-179c-484e-9f06-eaf8d27bca94/local.rds?X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Credential=ASIATLYQ5N5XYZBLLU6S%2F20221228%2Fus-west-2%2Fs3%2Faws4_request&X-Amz-Date=20221228T035446Z&X-Amz-Expires=604800&X-Amz-SignedHeaders=host&X-Amz-Security-Token=IQoJb3JpZ2luX2VjEPD%2F%2F%2F%2F%2F%2F%2F%2F%2F%2FwEaCXVzLXdlc3QtMiJHMEUCIBQJjSPuBtSseMrJG1qhgLJhbKTLNQbnlhXLQD9nYQ7fAiEAy3PBobg4vdpNo%2BDN%2FOmli9w4ilQr%2BMFu0%2BSBoN%2BAWnQq6wMIORABGgwyMzE0MjY4NDY1NzUiDLwAEyiaORP6sMM5VirIAwkrUFPpE0Jd1YsEvTcvuwhWEgV7bFInGPkh5I8SJ7OOJWr1lifuzO2pLqjtpZyfltJPKvzEI%2FWcM7H1yW4VNhDKTHARgEaR%2Fp4GscQ3Cs3XWXL8lTUtGHyhwLmiJLDhS%2BdqGehxj8AYpi8yxQvU6dEaq%2B2V3mMQBE6WOwoKDN%2FBPzCsXPldMvn3iJK07PKPLnwE%2FzER54G4KbMwdW%2FmoqeKuW1yU8MOv%2FHFwpbruxaF8NkoNcV24jtGJIzZTkTbbKkS%2Fr%2FM06AS3IpfS4veRV3SkDX4ZfqjNpIGjZTRseN%2FUzvtwmFD6Ubn%2FUEBOPKjhsKglB40rS7w2LBkakXefoTjbC%2BztQgnoj%2FT4tcmlmZzfOAoAsBGUqK4OEatYxeBAO2%2FVRpE%2FH9q%2Byrs1HkNWnK643TKnHLs6qE9bQWqq1ZULD4A9ElgkJPdZj64G9R4ZJuwZCuzXMEc2foeGwykzfQrj6h9qGzoVWMlDE01d0jCs9xxbYhG5yNkblf4q1YF2RzzREhc01FRHOlC4me6ghl9gUVefTISLrrUc9FfCm8Rk06UTPZueXKAoLafQFy9xqBrAKES9pXqSPV8FJsVndKpEtaA54SrozDui66dBjqlAd3Tm8NKevgOX5AukwiBH%2FqvrunrYbon1A%2BNhvcKHPnTU0P2eK%2FvOj1yUZbb%2FCQTbxuqAb3csEPB14XnaInQN3uramz%2BW9NCVDoXupLmb37MPQs%2BOpXF8jBL0PQLWzgAxzz9NCEYyRBeFmIRIRBfjqGVJQ3r%2F9I1NbrrSLTVziiZv0AP3jegAnz6mt68dnp%2BRCyJ%2BbRkFuzMFg%2B%2BMc88naeOvooPmg%3D%3D&X-Amz-Signature=e5039dae93e8ea7d5caa403195a6e1d1b975d601606be7b658e5ab1cbd16a820"`

```{r}
library(Seurat)
library(ggplot2)
```

## Load seurat dt

```{r}
dt = readRDS("dt/otero.rds")
```

dt is too big.. need to purge 

```{r}
summary(dt@meta.data$Cell.Types)
```
c("#ED2024", "#939598")

feature plot for PARP (ENSG00000143799)
```{r}
FeaturePlot(dt, features = "ENSG00000143799", cols = c("lightgrey","#ED2024")) 
ggsave("fig/PARP1_levels_UMAP.pdf", width = 6, height = 4)
```

feature plot for MTHFD2L ENSG00000163738

```{r}
FeaturePlot(dt, features = "ENSG00000163738", cols = c("lightgrey","#ED2024")) 
ggsave("fig/MTHFD2L_levels_UMAP.pdf", width = 6, height = 4)
```

```{r}
VlnPlot(object = dt, features = "ENSG00000143799", split.by = "disease")
```


```{r}
VlnPlot(object = dt, features = "ENSG00000163738", split.by = "disease")
```

### Exploratory analyses and QC 

Run this part in the cluster because too Long

Transformed data will be available in the SCT assay, which is set as the default after running sctransform

Formal class 'Seurat' [package "SeuratObject"] with 13 slots
  ..@ assays      :List of 1
  .. ..$ RNA:Formal class 'Assay' [package "SeuratObject"] with 8 slots
  .. .. .. ..@ counts       : num[0 , 0 ] 
  .. .. .. ..@ data
  
  

dt@assays$RNA@counts = dt@assays$RNA@data
dt@assays$RNA@data = new(Class = "matrix")
dt <- SCTransform(dt)
dt=RunPCA(dt, verbose = FALSE)
dt=RunUMAP(dt, dims = 1:30, verbose = FALSE)
dt <- FindNeighbors(dt, dims = 1:30, verbose = FALSE)
dt <- FindClusters(dt, verbose = FALSE)
saveRDS(dt, "norm_dt.rds")


Then, make a dataframe with the collowing columns: patientID, celltype, age, SCT_PARP1 
```{r}
# Visualize QC metrics as a violin plot
VlnPlot(dt, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```


```{r}
DimPlot(dt, reduction = "umap", group.by = "SORT")
```

```{r}
DimPlot(dt, reduction = "umap", group.by = "Amyloid")
```

```{r}
DimPlot(dt, reduction = "umap", group.by = "Cell.Types")
```

```{r}
DimPlot(dt, reduction = "umap", group.by = "disease")
```


## PARP1 levels

```{r} 
PARP1_dt = data.frame(cellID = dt@assays$RNA@data@Dimnames[2],
                        mt = dt@meta.data$percent.mt,
                        age = as.numeric(as.character(dt@meta.data$Age)),
                        type = dt@meta.data$Cell.Types,
                        braak = dt@meta.data$Braak,
                        disease = dt@meta.data$disease,
                      sex = dt@meta.data$sex,
                      ethnicity = dt@meta.data$self_reported_ethnicity,
                      patientID = dt@meta.data$donor_id,
                      sort = dt@meta.data$SORT,
                      PARP1 = GetAssayData(object = dt, slot = "data")["ENSG00000143799",])
PARP1_dt$logPARP1 = log2(PARP1_dt$PARP1)
write.csv(PARP1_dt, "dt_out/PARP1_dt_subset.csv", row.names = F)
```

```{r}
PARP1_dt_finite = subset(PARP1_dt, PARP1 != 0)
ggplot(PARP1_dt_finite, aes(x=(PARP1))) + geom_histogram()
```

```{r}
ggplot(PARP1_dt, aes(x=log2(PARP1))) + geom_histogram()
```

Data needs to be logged 

Initial model
```{r}
# ensure that the reference variable is "control patients" 
PARP1_dt_finite <- within(PARP1_dt_finite, disease <- relevel(disease, ref = 2))
PARP1_lm_simplified = lm(data = PARP1_dt_finite, formula = logPARP1 ~ disease + as.numeric(age) + sort +sex)
summary(PARP1_lm_simplified)
```

(Intercept)               0.354729   0.035003   10.13   <2e-16 ***
diseaseAlzheimer disease  0.469339   0.009243   50.78   <2e-16 ***
as.numeric(age)          -0.009765   0.000475  -20.56   <2e-16 ***
sexmale                  -0.012811   0.008596   -1.49    0.136    

Note: lmerTest uses the stringent Satterthwaite approximation, which is based on SAS proc mixed theory.

```{r}
library(lmerTest)
PARP1_lm_simplified_lmer = lmer(data = PARP1_dt_finite, formula = logPARP1 ~ disease + as.numeric(age) + sex+sort+ (1 | patientID), REML = F)
summary(PARP1_lm_simplified_lmer)

```

                          Estimate Std. Error        df t value Pr(>|t|)  
(Intercept)               0.159887   0.540421 16.010590   0.296   0.7711  
diseaseAlzheimer disease  0.396011   0.140749 16.006228   2.814   0.0125 *
as.numeric(age)          -0.006067   0.007132 16.014924  -0.851   0.4075  
sexmale                   0.051760   0.142128 16.001549   0.364   0.7205  

```{r}
yy_plt_forest_lm = function(lm_model){
  coef_mat = as.data.frame(summary(lm_model)$coefficients)
  coef_mat = na.omit(coef_mat)
  conf = na.omit(confint.default(lm_model, level = 0.95))
  coef_mat$lci = conf[,1]
  coef_mat$uci = conf[,2]
  coef_mat$label = row.names(coef_mat)
  coef_mat = subset(coef_mat, label != "(Intercept)")
  ggplot(data=coef_mat, aes(x=reorder(label,Estimate),y=Estimate, fill = `Pr(>|t|)` < 0.05, color = `Pr(>|t|)` < 0.05)) +
    geom_hline(yintercept=0, lty=2) +  
    geom_errorbar(aes(ymin=lci, ymax=uci),
                  width=0,                    # Width of the error bars
                  position=position_dodge(.9), color = "grey", size = 1.5) +
    geom_point(shape=21, size = 2.5)+
    #geom_pointrange(aes(fill = `Pr(>|t|)` < 0.05)) +
    coord_flip() +  # flip coordinates (puts labels on y axis)
    xlab("") + ylab("Effect (95% CI)") +
    scale_fill_manual(values = c("#ED2024", "#939598"))+
    scale_color_manual(values = c("#ED2024", "#939598"))+
    theme_classic()+
    theme(axis.text.x=element_text(colour="black"),
          axis.text.y=element_text(colour="black"))
}
yy_plt_forest_lm(PARP1_lm_simplified)
ggsave("fig/PARP1_single_cell_expression.pdf", width = 5, height = 2)
```


Add more covariates
```{r}
PARP1_lm = lm(data = PARP1_dt_finite, formula = logPARP1 ~ disease + as.numeric(age) + mt + type + braak + sex + ethnicity)
summary(PARP1_lm)
yy_plt_forest_lm(PARP1_lm)
ggsave("fig/PARP1_single_cell_expression_full_model.pdf", width = 5, height = 3)
```

```{r}
PARP1_lm_full_lmer = lmer(data = PARP1_dt_finite, formula = logPARP1 ~ disease + as.numeric(age) + mt + type + braak + sex + ethnicity+ (1 | patientID), REML = F)
summary(PARP1_lm_full_lmer)
```

Check the reverse
```{r}
AD_PARP1_glm_simplified = glm(data = PARP1_dt_finite, formula = disease ~ logPARP1 + as.numeric(age) + sex + ethnicity, family = "binomial")
summary(AD_PARP1_glm_simplified)
```

```{r}
AD_PARP1_glm = glm(data = PARP1_dt_finite, formula = disease ~ logPARP1 + as.numeric(age) + sex + ethnicity + mt + type, family = "binomial")
summary(AD_PARP1_glm)
```

diseaseAlzheimer disease             0.2822209  0.0147432  19.142  < 2e-16 ***
as.numeric(age)                     -0.0048049  0.0013197  -3.641 0.000273 ***

```{r}
ggplot(PARP1_dt_finite, aes(x=as.numeric(age), y=logPARP1, color=disease, fill = disease)) +
  geom_point(position = position_jitter(seed = 1, width = 0.2),
             alpha = 0.5)+
  geom_smooth(method = "lm")+
  scale_fill_manual(values=c("#ED2024", "#939598"))+
  scale_color_manual(values=c("#ED2024", "#939598"))+
  theme_classic()
```

```{r}
data_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

ggplot(PARP1_dt_finite, aes(x=disease, y=logPARP1, color=disease)) +
    geom_violin(trim = F) + 
  #geom_point(position = position_jitter(seed = 1, width = 0.2), alpha = 0.5)+
  stat_summary(fun.data=data_summary)+
  scale_fill_manual(values=c("#ED2024", "#939598"))+
  scale_color_manual(values=c("#ED2024", "#939598"))+
  theme_classic()
ggsave("fig/PARP1_expression_violin.pdf", width = 4, height = 3)
```

Add cell numbers

```{r}
summary(PARP1_dt_finite)
```
normal           :10390
Alzheimer disease: 8013



