---
title: "Analysis of MTHFD2L mRNA levels in scRNA seq"
author: "Yizhou Yu"
date: 'updated: <i>`r format(Sys.Date())`</i></h4>'
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
    keep_md: yes
---
download data: 

`curl -o local.rds "https://corpora-data-prod.s3.amazonaws.com/f541adea-179c-484e-9f06-eaf8d27bca94/local.rds?X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Credential=ASIATLYQ5N5XYZBLLU6S%2F20221228%2Fus-west-2%2Fs3%2Faws4_request&X-Amz-Date=20221228T035446Z&X-Amz-Expires=604800&X-Amz-SignedHeaders=host&X-Amz-Security-Token=IQoJb3JpZ2luX2VjEPD%2F%2F%2F%2F%2F%2F%2F%2F%2F%2FwEaCXVzLXdlc3QtMiJHMEUCIBQJjSPuBtSseMrJG1qhgLJhbKTLNQbnlhXLQD9nYQ7fAiEAy3PBobg4vdpNo%2BDN%2FOmli9w4ilQr%2BMFu0%2BSBoN%2BAWnQq6wMIORABGgwyMzE0MjY4NDY1NzUiDLwAEyiaORP6sMM5VirIAwkrUFPpE0Jd1YsEvTcvuwhWEgV7bFInGPkh5I8SJ7OOJWr1lifuzO2pLqjtpZyfltJPKvzEI%2FWcM7H1yW4VNhDKTHARgEaR%2Fp4GscQ3Cs3XWXL8lTUtGHyhwLmiJLDhS%2BdqGehxj8AYpi8yxQvU6dEaq%2B2V3mMQBE6WOwoKDN%2FBPzCsXPldMvn3iJK07PKPLnwE%2FzER54G4KbMwdW%2FmoqeKuW1yU8MOv%2FHFwpbruxaF8NkoNcV24jtGJIzZTkTbbKkS%2Fr%2FM06AS3IpfS4veRV3SkDX4ZfqjNpIGjZTRseN%2FUzvtwmFD6Ubn%2FUEBOPKjhsKglB40rS7w2LBkakXefoTjbC%2BztQgnoj%2FT4tcmlmZzfOAoAsBGUqK4OEatYxeBAO2%2FVRpE%2FH9q%2Byrs1HkNWnK643TKnHLs6qE9bQWqq1ZULD4A9ElgkJPdZj64G9R4ZJuwZCuzXMEc2foeGwykzfQrj6h9qGzoVWMlDE01d0jCs9xxbYhG5yNkblf4q1YF2RzzREhc01FRHOlC4me6ghl9gUVefTISLrrUc9FfCm8Rk06UTPZueXKAoLafQFy9xqBrAKES9pXqSPV8FJsVndKpEtaA54SrozDui66dBjqlAd3Tm8NKevgOX5AukwiBH%2FqvrunrYbon1A%2BNhvcKHPnTU0P2eK%2FvOj1yUZbb%2FCQTbxuqAb3csEPB14XnaInQN3uramz%2BW9NCVDoXupLmb37MPQs%2BOpXF8jBL0PQLWzgAxzz9NCEYyRBeFmIRIRBfjqGVJQ3r%2F9I1NbrrSLTVziiZv0AP3jegAnz6mt68dnp%2BRCyJ%2BbRkFuzMFg%2B%2BMc88naeOvooPmg%3D%3D&X-Amz-Signature=e5039dae93e8ea7d5caa403195a6e1d1b975d601606be7b658e5ab1cbd16a820"`


```{r}
library(Seurat)
library(ggplot2)
```

## Load seurat dt

```{r}
dt = readRDS("dt/otero.rds")
```

dt is too big.. need to purge 

feature plot for MTHFD2L ENSG00000163738

```{r}
FeaturePlot(dt, features = "ENSG00000163738", cols = c("lightgrey","#ED2024")) 
ggsave("fig/MTHFD2L_levels_UMAP.pdf", width = 6, height = 4)
```

```{r}
VlnPlot(object = dt, features = "ENSG00000163738", split.by = "disease")
```

### Exploratory analyses and QC 


## PARP1 levels

```{r} 
MTHFD2L_dt = data.frame(cellID = dt@assays$RNA@data@Dimnames[2],
                        mt = dt@meta.data$percent.mt,
                        age = as.numeric(as.character(dt@meta.data$Age)),
                        type = dt@meta.data$Cell.Types,
                        braak = dt@meta.data$Braak,
                        disease = dt@meta.data$disease,
                      sex = dt@meta.data$sex,
                      ethnicity = dt@meta.data$self_reported_ethnicity,
                      patientID = dt@meta.data$donor_id,
                      sort = dt@meta.data$SORT,
                      Exp = GetAssayData(object = dt, slot = "data")["ENSG00000163738",])
MTHFD2L_dt$logExp = log2(MTHFD2L_dt$Exp)
write.csv(MTHFD2L_dt, "dt_out/MTHFD2L_dt_subset.csv", row.names = F)
```

```{r}
MTHFD2L_dt_finite = subset(MTHFD2L_dt, Exp != 0)
ggplot(MTHFD2L_dt_finite, aes(x=(logExp))) + geom_histogram()
```

Data needs to be logged 

Initial model
```{r}
# ensure that the reference variable is "control patients" 
MTHFD2L_dt_finite <- within(MTHFD2L_dt_finite, disease <- relevel(disease, ref = 2))
lm_simplified = lm(data = MTHFD2L_dt_finite, formula = logExp ~ disease + as.numeric(age) + sex)
summary(lm_simplified)
```

                          Estimate Std. Error t value Pr(>|t|)    
(Intercept)               0.1490257  0.0327658   4.548 5.44e-06 ***
diseaseAlzheimer disease  0.4066557  0.0087456  46.498  < 2e-16 ***
as.numeric(age)          -0.0056267  0.0004398 -12.794  < 2e-16 ***
sexmale                   0.0238828  0.0081157   2.943  0.00326 ** 

Note: lmerTest uses the stringent Satterthwaite approximation, which is based on SAS proc mixed theory.

```{r}
library(lmerTest)
lm_simplified_lmer = lmer(data = MTHFD2L_dt_finite, formula = logExp ~ disease + as.numeric(age) + sex+ (1 | patientID), REML = F)
summary(lm_simplified_lmer)

```

                          Estimate Std. Error        df t value Pr(>|t|)   
(Intercept)               0.203991   0.463987 16.016388   0.440  0.66607   
diseaseAlzheimer disease  0.367286   0.120857 16.020265   3.039  0.00781 **
as.numeric(age)          -0.005529   0.006124 16.021188  -0.903  0.37993   
sexmale                   0.063264   0.122019 16.003467   0.518  0.61122   

```{r}
yy_plt_forest_lm = function(lm_model){
  coef_mat = as.data.frame(summary(lm_model)$coefficients)
  coef_mat = na.omit(coef_mat)
  conf = na.omit(confint.default(lm_model, level = 0.95))
  coef_mat$lci = conf[,1]
  coef_mat$uci = conf[,2]
  coef_mat$label = row.names(coef_mat)
  coef_mat = subset(coef_mat, label != "(Intercept)")
  ggplot(data=coef_mat, aes(x=reorder(label,Estimate),y=Estimate, fill = `Pr(>|t|)` < 0.05, color = `Pr(>|t|)` < 0.05)) +
    geom_hline(yintercept=0, lty=2) +  
    geom_errorbar(aes(ymin=lci, ymax=uci),
                  width=0,                    # Width of the error bars
                  position=position_dodge(.9), color = "grey", size = 1.5) +
    geom_point(shape=21, size = 2.5)+
    #geom_pointrange(aes(fill = `Pr(>|t|)` < 0.05)) +
    coord_flip() +  # flip coordinates (puts labels on y axis)
    xlab("") + ylab("Effect (95% CI)") +
    scale_fill_manual(values = c("#ED2024", "#939598"))+
    scale_color_manual(values = c("#ED2024", "#939598"))+
    theme_classic()+
    theme(axis.text.x=element_text(colour="black"),
          axis.text.y=element_text(colour="black"))
}
yy_plt_forest_lm(lm_simplified)
ggsave("fig/MTHFD2L_single_cell_expression.pdf", width = 5, height = 2)
```


Add more covariates
```{r}
lm = lm(data = MTHFD2L_dt_finite, formula = logExp ~ disease + as.numeric(age) + mt + type + braak + sex + ethnicity)
summary(lm)
yy_plt_forest_lm(lm)
ggsave("fig/MTHFD2L_single_cell_expression_full_model.pdf", width = 5, height = 3)
```

```{r}
lm_full_lmer = lmer(data = MTHFD2L_dt_finite, formula = logExp ~ disease + as.numeric(age) + mt + type + braak + sex + ethnicity+ (1 | patientID), REML = F)
summary(lm_full_lmer)
```

```{r}
ggplot(MTHFD2L_dt_finite, aes(x=as.numeric(age), y=logExp, color=disease, fill = disease)) +
  geom_point(position = position_jitter(seed = 1, width = 0.2),
             alpha = 0.5)+
  geom_smooth(method = "lm")+
  scale_fill_manual(values=c("#939598","#ED2024"))+
  scale_color_manual(values=c("#939598","#ED2024"))+
  theme_classic()
ggsave("fig/MTHFD2L_single_cell_scatter_age_disease.pdf", width = 5, height = 3)
```

```{r}
data_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

ggplot(MTHFD2L_dt_finite, aes(x=disease, y=logExp, color=disease)) +
    geom_violin(trim = F) + 
  #geom_point(position = position_jitter(seed = 1, width = 0.2), alpha = 0.5)+
  stat_summary(fun.data=data_summary)+
  scale_fill_manual(values=c("#939598","#ED2024"))+
  scale_color_manual(values=c("#939598","#ED2024"))+
  theme_classic()
ggsave("fig/MTHFD2L_expression_violin.pdf", width = 4, height = 3)
```

Add cell numbers

```{r}
summary(MTHFD2L_dt_finite)
```
normal           :12167
Alzheimer disease: 9173



